---
title: "Spectra reconstruction analysis"
author: "Ellie Dunstone"
date: "2023-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(tidyverse)
library(patchwork)
library(sigfit)
```


## Introduction

This is a script for analysing and exploring the quality of the spectrum reconstructions generated during signature extraction or fitting using sigfit, which can be installed from <https://github.com/kgori/sigfit>. Reconstructions of your samples under the 'model' you have calculated can be produced for all three modalities implemented in sigfit: reference signature fitting; de novo signature extraction; and combined fitting and extraction (FitExt). For further explanation, see the vignettes on the sigfit github, or my own 'sigfit_signature_analysis.Rmd' script that implements the sigfit package.

Sigfit allows us to examine how effectively the estimated signatures and/or exposures can reconstruct the original mutational catalogues by producing spectrum reconstruction plots with the 'plot_reconstruction' function. The underlying data can be accessed using the 'get_reconstructions' function, which provides the input data for this script. Some of the content of this script is adapted from the 'plot_reconstruction' function written by the authors of sigfit.

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
